% R Introduction for UCL PhDs
% UCL 2013
% Florian Oswald

# Welcome to R!
## Contents
>1. Some basic R. Not (at all) exaustive. 
    * These slides are designed to get you started. 
    * Try this [online R challenge at codeschool](http://tryr.codeschool.com/levels/1/challenges/1). Codeschool provides excellent tryouts IN YOUR BROWSER (you don't need to install anything.
    * Have a look at many more resources about learning R on [my UCL wiki](https://wiki.ucl.ac.uk/display/~uctpfos/R-Resources) 
>2. Some quizzes
>3. Some example usage. We'll be scratching the surface.
>4. A sample applied econometric project using the FES: reading, merging, analysing, transforming data, plots and regressions
>5. R and foreign Languages
>6. How to make this set of slides. 

# Getting R and RStudio
>* [Click here for www.r-project.org and download R for your OS](http://www.r-project.org)
>* [Click here for www.rstudio.org to download RStudio for your OS](http://www.rstudio.org)
>* **Take a moment to contemplate what you just did**
    * you downloaded a programme from the internet without paying
    * what you did was completely legal
>* [As Roger Koenker would say](http://www.econ.uiuc.edu/~roger/courses/LSE/outline/yaRfaq.pdf): *Buy a case of wine with what you've saved*

# Why R?

>* Licensing
>* Availability of add on software (packages)
>* Very easy to adapt to your needs. You can look at existing code, change it and reuse it.
>* Ease of connecting to other (faster) software like fortran or C
>* Who uses R? (Apart from Roger Koenker?)
    * Very active and growing community in science and business.
    * data scientists use R
    * [who in business uses R?](http://www.revolutionanalytics.com/what-is-open-source-r/companies-using-r.php) 

# Why not R?

* steep learning curve. 
* "R is slow". Although this is a very imprecise statement. See "calling C and Fortran".
* the sheer extent of packages available makes is hard at time to find the most suitable one
* Some people lament that the CRAN model led to a *bastardization* of the language, and that it should be rewritten from scratch. 

# Learning R if you know Matlab/Stata/other

* [R for Stata Users](http://dss.princeton.edu/training/RStata.pdf)
* [R for Matlab Users](http://cran.r-project.org/doc/contrib/Hiebeler-matlabR.pdf)
* more links on my wiki...
* get familiar with the R community. Checkout 
    * [Stackoverflow](http://www.stackoverflow.com)
    * [R mailing lists on many topics](http://www.r-project.org/mail.html)

# Supplementary material 

* We'll use some sample FES data and a Fortran function at some point
* if you are a git user type in your terminal 

````
git clone git://github.com/floswald/R-demo.git
````

* otherwise go to [https://github.com/floswald/R-demo](https://github.com/floswald/R-demo) and click on the button with the cloud and "ZIP" on it

# Using this set of slides

* hit `c` on your keyboard for a linked table of contents. `c` again makes it disappear.
* left/right or mouse click or swipe takes you around
* code looks like this:
```{r}
x <- 3.4
x  # print x to screen
```
* the first block is *input*, the second block (marked with `##`) is output
* everything after `#` is a comment
* you can (and should) type or copy and paste the code into your R and try around. All of the code you'll see actually works.
* let's do it.

# Getting `help()`

* if you want help about a function, type `help(fname)`
* if you don't know which function you are looking for, try `??fname``
* go to the [R Site Search](http://finzi.psych.upenn.edu/search.html)

# Basics

* Case-sensitive: x is different from X
* fully-fledged programming language. all the usual flow controls
    * `if`, `for`, `while`, etc
    * [check out the language definition](http://cran.r-project.org/doc/manuals/R-lang.html) if needed
* R needs to have all objects in memory (there are exeptions)
* R's memory is more like matlabs and less like stata's
    * you can have many objects of different size, type etc
    * (stata's memory is one big spreadsheet. you can store a vector on the side, but it's complicated.)



# Assigning Values to Objects

- Everything in R (numbers, letters, vectors, functions, plots, formulas, calls,...) is an object
- assign a value (a number, a matrix, a word, a function, ...) to an object with ```<-``` or ```=``` . Use ```<-``` .
```{r assignx}
x <- 3.4
x
```
reassigning overwrites the current value:
```{r}
x <- "now x is a character string."
x
x <- c(3,5,2.1,1001,4.6)  # c() function combines single values into a vector. 
x
```

# The functions ```c()```, ```str(x)``` and ```typeof(x)```

* ```typeof(x)``` tells you how x was stored, ```mode(x)``` is similar
* among possible storage types are
    * ```logical```: TRUE or FALSE
    * ```integer```: 1, 2, 3, -10, ...
    * ```double```: 1.0, pi, -3.2331, ...
    * ```factor```: categorial variables
        * dummy vars: (0,1) labelled `male`,`female`
	      * categories: (1,2,3) labelled `(2,5],(5,7],(7,10]`	
    * ```complex```: 0+0i
    * ```character```: "hello world"
* ```c()``` requires values to have the same type (or it coerces into a common type)
```{r coerce}
typeof(x)
y <- letters[10:20]   # letters number 10 trough 20 of the english alphabet
typeof(y)
typeof(c(x,y))
str(c(x,y))   # compact display of structure of an object
```

# Subsetting a vector

* access the n-th element of ```x``` with ```x[n]```
* access any tuple of elements by suppling the indices
```{r subsetx}
x <- rnorm(n=8) # draw 8 random normal values
x
x[3:6] # "3:6" produces c(3,4,5,6) 
x[c(1,5,8)] # elts 1,5 and 8
x[-c(1,5,8)] # all elts except 1,5 and 8
```

# Helper Functions for allocation: ```seq()``` and ```rep()```

```{r helpers}
x <- seq(from=1,to=15,by=3) # also "length" instead of "by". like matlab linspace()
x
str(x)
y <- rep(1:3,c(2,3,4))
y
z <- rep(c("oh my word"),3)
z
```

# Your turn 1

* create a vector x with values 0, 3,  6,  9, 12, 15, 18
* what is the ```typeof(x)```?
* create a vector y containing the words {"are","my","favourite","numbers"}
* create a vector z from the last 2 elements of x and all of y
* what's the ```typeof(z)```?

# Your solution 1

```{r ysol1}
x <- seq(from=0,to=18,by=3)
x
typeof(x)
y <- c("are","my","favourite","numbers")
z <- c(x[c(6,7)],y)	# R "coerces" x and y to a common type. here: char
z
typeof(z)
```


# Some Vector Arithmetic

* Operations +-/* are element by element
```{r arith}
x <- 1:3
y <- 4:6
x + y
```
* and vectors are "recycled" if they are not of the same length
```{r recycle}
x <- 1:4
x + y  # recycling the shorter vector y. R gives a warning.
```

# matrix
```{r matrix}
m1 <- matrix(data=1:9,nrow=3,ncol=3,byrow=TRUE)
m2 <- matrix(data=1:12,nrow=4,ncol=3,byrow=FALSE)
m1
m2
```

# Combining matrices

* similarly to ```c()``` for vectors, we have ```cbind()``` and ```rbind()```
* column-bind and row-bind
```{r matrix2}
rbind(m1,m2) # glue together the last row of of m1 and first of m2
cbind(m1,t(m2))  # glue last col of m1 and first of t(m2)
```


# Subsetting a matrix
* get size of matrix with ```dim(m1)```
* you can select indices of rows and columns. Very similar to matlab.
```{r submat}
m1[2, ]   # row 2, all columns
m2[ ,1]   # all rows, column 1
m1[c(1,3),c(2,3)] # rows 1,3 and cols 2,3
```
* or by name, if those were given:
```{r submat2}
colnames(m1) <- c("col1","col2","col3")
m1[,"col2"]
colnames(m1) <- NULL # remove colnames
```

# basic matrix arithmetic
Again, +-/* are element-wise. Works only on equal sized matrices:
```{r mat1}
m1 + m2  # error
m1 * m2  # error
```
```%*%``` is the matrix muliplication operation. matrices need to be conformable:
```{r mat2}
m2 %*% m1
m1 %*% t(m2)
m1 %*% m2
```

# data.frame
* A data.frame is a standard dataset
* very similar to ```matrix```
* each row is an observation, each column is a variable
* create a data.frame from scratch with ```data.frame()```
* Notice: all columns need to be of the same length (or *recyclable*)! (a data.frame is a rectangle; it's a spreadsheet!)
```{r df}
df <- data.frame(cat.1=rep(1:3,each=2),cat.2=1:2,values=rnorm(6))
dim(df)
df
```

# Subsetting a data.frame

* A data.frame is similar to a matrix. Subsetting is thus similar.
* but additionally, you can access columns with the ```$``` operator:
```{r subdf}
df$values
```
* the ```$``` in general accesses elements of "lists" (see below)
* if an element does not exist, ```$``` adds it:
```{r subdf2}
df$new.col <- df$cat.1 + df$values  
df
```

# ```subset()``` 1

* get a subset of observations conditional on values of a variable
```{r subdf3}
subset(df,cat.1==1) # select all obs were cat.1 equals 1
```
* this is the same as ```df[df$cat.1==1,]```

# ```subset()``` 2

* get a subset of variables 
```{r subdf4}
subset(df,select = c(cat.2, values)) # selects columns "cat.2" and "values"
```

# Removing columns

remove a column with ```NULL```:
```{r subdf5}
df$new.col <- NULL  
df
```

# Appending Data: rbind

* very similar to ```rbind``` for ```matrix```
* If you have two data.frames with equally named columns (e.g. two editions of the same dataset), just glue together with ```rbind```

```{r subdf6}
df2 <- df # create df2 as an exact copy of df
df2$values <- 1:nrow(df)  # but change the entries for 'values'
rbind(df,df2) # join them row-wise
```

# Looking at data.frames
* R stores datasets typically as data.frames
* special "methods"" to describe those (a method is a function that does things differently depending on the object you give it)
```{r data1}
data(LifeCycleSavings) # data() shows available built-in datasets
LS <- LifeCycleSavings
head(LS)	# show the first 6 rows of data.frame LifecycleSavings.
summary(LS) # function summary()
```

# ordering a data.frame

* rearrange row indices by a certain criterion
* ```order()``` and ```sort()```
* function order(x) returns the indices of x in ascending/descending order:
```{r}
order(c(5,2,6,1),decreasing=TRUE)
```
* plugging in a vector of indices into a matrix/data.frame, reorders the matrix according to those:
```{r order}
save.ranking <- order(LS$sr,decreasing=TRUE)	
head(LS[save.ranking,])
```

# packages
* "packages" are add on libraries, i.e. collections of functions, algorithms and datasets
* The great number of available packages is one of the strengths of R. 
* packages on CRAN have a certain degree of quality control

# finding packages

* googling what you want to do often helps
* for exampe, [let me google "R optimization" for you](http://lmgtfy.com/?q=R+optimization)


# installing packages

* install a packages available at CRAN simply by typing, for example ```install.packages("ggplot2")```
* `ggplot2` is a widely used package with plotting functions. 
* load a package by calling the library function:

```{r package}
library(ggplot2) # now the content of ggplot2 (functions, data, etc) are "visible"
cut_interval     # for example, here's the code to function "cut_interval"
```
* get help for a package via ```help(package="package-name")```

# Functions

* R relies heavily on functions
* in fact, R is composed of functions
* look at the source code of any R function by typing the name:
```{r func1}
matrix
```
* you can of course define your own functions.
* for example, a function to compute the interval that is "spanned" by a vector, i.e. the interval between the min and max of vector:
```{r func2}
span <- function(x) {
  stopifnot(is.numeric(x))  # stops if x is not numeric
  r <- range(x)   # range() gives the range
  rval <- abs(r[2]-r[1])  # computes and returns the interval spanned by x
  return(rval)  # returns result
  }
myvec <- rnorm(50)   # draws 50 values from the standard normal pdf
span(myvec)
```

# Lists

* can contain data in any "mode": vectors, matrices, data.frames, characters, functions, plots, and more lists
* similar to a `matlab` or `C` *structure* 
```{r list1}
l <- list(words=c("oh my word(s)"),mats=list(mat1=m1,mat2=m2),funs=span)
l
```

# Working with Lists

* Access a list element by position number in the list, or by name (if given)
* ```[["x"]]``` gives you element "x" as a standalone object
* ```["x"]``` gives you element "x" as a sublist with element "x"
```{r list2}
l[[2]]  # second list element: "mats"
l$mats$mat1 	# of that element, the first element
str(l[[1]])	 # extract first element completely
str(l[1])	 # extract first element as member of sublist
```

# Adding and removing a new elements from a list

```{r list3}
l$new.element <- rnorm(5)    # add a new element: 5 random draws from the standard normal
l$bool.value <- l$new.element>0   # add a new element: previously drawn numbers positive?
l$new.element <- NULL   # delete an element
l$bool.value <- NULL
l$words <- NULL
l$funs <- NULL
l
```

# Your Turn 2

* make a list "fruit" with two elements: "apples" and "pears". 
* for both apples and pears make up 3 pairs of random price/quantity values, e.g. for apples    

 observation    price     quantity
-------------  --------  ----------
    1             1.3        2
    2             4.8        5
    3             8.9        10

* for both apples and pears, store those values (i.e. the above table) in a data.frame with colum names "price" and "quantity"
* In other words, produce the following list. (make up your own numbers)
```{r sol3,echo=FALSE}
fruit <- list(apples=data.frame(price=runif(min=0,max=4,n=3),quantity=sample(1:4,size=3)),pears=data.frame(price=runif(min=0,max=4,n=3),quantity=sample(1:4,size=3)))
fruit
```

# Your Solution 2

* you can input the values by hand, for example 

```
apples <- data.frame(price=c(1.1,2.9,1.6),quantity=c(2,3,1))
```

* I used the uniform random generator ```runif()``` for prices and ```sample()``` for quantities

```{r sol4}
apples <- data.frame(price=runif(min=0,max=4,n=3),quantity=sample(1:4,size=3))
pears <- data.frame(price=runif(min=0,max=4,n=3),quantity=sample(1:4,size=3))
fruit <- list()
fruit$apples <- apples
fruit$pears <- pears
# or all in one line:
fruit <- list(apples=data.frame(price=runif(min=0,max=4,n=3),quantity=sample(1:4,size=3)),pears=data.frame(price=runif(min=0,max=4,n=3),quantity=sample(1:4,size=3)))
```

# Factors

* Factors are categorical variables
* a number is given a character label
* R has plotting and tabulating methods for factors
* also nice useage in regressions
```{r}
new.fac <- factor(x=c(1,1,2,5,5,5),labels=c("apples","pears","shoes"))
new.fac
table(new.fac)
plot(new.fac)
```

# Contingency Tables of multiple Factors
* `table()` can count cases of factors against each other
```{r}
UCB <- as.data.frame(UCBAdmissions) 	# load built-in data on Berkeley admissions
tail(UCB) # this time, display the last 6 rows
summary(UCB)	# notice how summary treats factors and numerics
table(UCB$Admit,UCB$Dept) 	# table admit vs Dept?
```
* But is that what we want here?

# Contingency tables 2
* No! We want to know "Freq" by those categories (those ```factors```).
* In R-speak, what is the relationship ```Freq ~ factors```?
* `xtabs()`
  * takes a ```formula``` as an argument
  * typical formula: ```y ~ x + z``` means "y is explained by x and z"
```{r}
xtabs(Freq ~ Admit + Dept,data=UCB)	# Freq is explained by Admit and Dept
xtabs(Freq ~ Admit + Gender,data=UCB)
```

# Your turn 3
* create a table that shows the number of admissions/rejections by gender AND department.

# Your Solution 3
```{r}
xtabs(Freq ~ Admit + Gender + Dept,data=UCB)
```


# Workspace
* In general, R needs to load an object into memory to be able to work with it
* This can be a problem with large data sets
* the function ```ls()``` shows the content of your workspace, or your current "search path"
```{r ls}
ls()  
rm(df,m1,m2,"%ni%",fruit,l)	# remove objects
ls()
rm(list=ls(all=TRUE))	# remove all objects
ls()
```

# A Sample Applied Econometric Project

## Food and Expenditure Survey (FES)

* the repository you downloaded from github contains a folder "data"
* contains 4 files, 2 in .csv and 2 in .dta format
* each format has one data file ("fesdat") and one file on income ("fesinc")

## Agenda

1. how to read different data
2. how to merge data
3. how to analyse data
    * with summary statistics (numerical variables)
    * in tables (categorical vars)
    * graphically
    * with statistical models
4. how to report results

# R does a lot of Econometrics 
* [click here to get an overview of the Econometrics capabilities of R.](http://cran.stat.ucla.edu/web/views/Econometrics.html)

# Reading Data Files

* Read data with with family of ```read.xxx()``` functions: 
  * ```read.csv()```: comma separated values
  * ```read.dta()```: stata format
  * ```read.table()```: generic text/table reader
* ```library(foreign)``` contains many useful functions to read all sorts of foreign data formats
* ```library(xlsx)``` is useful to read .xls data

# Read FES

* let's read both .csv and .dta formats just to demonstrate:
```{r}
setwd(dir="~/git/R-demo/")   # set working directory
fesdat.csv <- read.csv(file="data/fesdat.csv") # read the data in csv format. Note that "file" could also be a URL
fesinc.csv <- read.csv(file="data/fesinc.csv") # read the income data in csv format
library(foreign)  # load foreign to read stata data
fesdat.dta <- read.dta(file="data/fesdat2005.dta") 
fesinc.dta <- read.dta(file="data/fesinc2005.dta") 
head(fesdat.dta) # do they look the same?
head(fesdat.csv)
```

# Summary

```{r}
dat <- fesdat.dta # let's rename this to something shorter
inc <- fesinc.dta
summary(dat) 
dat$year <- NULL # we've only got one year, so useless info
```

# Factors
```{r}
str(dat$marstat) # marstat is already a factor
levels(dat$marstat)
dat$sex <- factor(dat$sex,labels=c("male","female"))  # convert sex into a factor
summary(subset(dat, select = c(sex,marstat)))
```

# Contingency Tables of Factors

```{r}
attach(dat) # attach(x) makes all cols of x visible on the search path
table(sex) # otherwise I'd have to type dat$sex here
table(marstat,sex)
```

# More complicated tables: ```xtabs()```

```{r}
f <- ~ sex + kids0
class(f)  # formula with only RHS "explaining" variables: just get a count
xtabs(f)  
xtabs(~ kids0 + sex + nrooms)	# three way table
prop.table(xtabs(~ sex + kids0),margin=1)	# proportion of kids0 by sex
prop.table(xtabs(~ sex + kids0),margin=2) # proportion of kids0 within age
detach(dat) # remove dat from search path
```

# Merge Data

* Both `dat` and `inc` include a unique hh identifier `hhref`
* use `merge()`
```{r}
dat.inc <- merge(dat,inc,by=c("hhref"))  # if multiple keys: by=c("hhref","year")
head(dat.inc) # note new column hhinc
```



# Create Bins: ```cut()``` an interval

* suppose instead of ```age = {17,18,19,...}``` we want categories of age
* base R has ```cut()```
* ggplot2 has derived ```cut_interval()```, bit more intuitive

```{r,message=FALSE}
dat.inc$age.cat <- cut_interval(dat.inc$age,length=5) # add col with age category
dat.inc$ced.cat <- cut_interval(dat.inc$ageced,length=5) # add col with age at educ completion category
str(dat.inc$age.cat) # is a factor?
table(dat.inc$age.cat)
plot(dat.inc$age.cat)
table(dat.inc$ced.cat)
```

# Create Bins 2: decile of income?

* Suppose we want to put people into their respective income decile
```{r}
inc.dec <- quantile(dat.inc$hhinc,prob=seq(0,1,length=11)) # income deciles
inc.dec
dat.inc$inc.cat <- cut(dat.inc$hhinc,inc.dec,labels=FALSE)
head(dat.inc[,c("hhinc","inc.cat")])	# look at columns hhinc and inc.cat
table(dat.inc$inc.cat)
```






# Summaries by groups of Variables: ```library(plyr)```
* Often we want summary statistics by groups of variables
* ```tapply()```, ```by()``` and ```aggregate()``` are base R possibilities
* we'll show ```library(plyr)``` here
* [click for the amazing tutorial of plyr by Hadley Wickham](http://plyr.had.co.nz/09-user/)
* workhorse functions are 
  * ```tranform()```: transform a data.frame, (you get a df of same or bigger size)
  * ```summarise()```: summarises a data.frame (you get a df that is smaller)
  
# ```summarise()``` and ```transform()```
* use our FES data in ```dat.inc```.
* functions work on an entire data.frame
```{r demoplyr,message=FALSE}
library(plyr)
# summarise an entire data.frame
summarise(dat.inc,mhhinc = mean(hhinc), vhhinc = sd(hhinc), mnumads=mean(numads)) 
# add "income rank" to dat.inc. head(...,10) to see first 10 lines only
head(transform(dat.inc,inc.rank = rank(hhinc)),10) 
```

# ```summarise()``` and ```transform()``` by groups?

1. how can we get the income rank for all categories of, say, ```marstat```?
2. can we express everybody's income as a percentage of the highest income in each category of ```marstat```?
3. what is the median ratio of durable to nondurable expenditure by number of rooms? 
4. what is it by incoem decile?
5. what is it by both?

# ```transform()``` by groups: ```ddply()```

1. add income rank for each household for all categories of ```marstat``` to ```dat```
```{r}
head(ddply(.data=dat.inc,.variables="marstat",.fun=transform, rank.marstat = rank(hhinc)),10)
```
2. express everybody's income as a percentage of the highest income in each category of ```marstat```
```{r}
head(ddply(dat.inc,"marstat",transform, perc.of.max = hhinc/max(hhinc)),10)
```

# ```summarise()``` by groups

3. median ratio of durable to nondurable expenditure by number of rooms 
```{r}
ddply(dat.inc,c("nrooms"),summarise, med.nondur.dur = median(nondur/durables,na.rm=TRUE))
```
4. by income decile
```{r}
ddply(dat.inc,c("inc.cat"),summarise, med.nondur.dur = median(nondur/durables,na.rm=T))
```
5. by both
```{r}
ddply(dat.inc,c("nrooms","inc.cat"),summarise, med.nondur.dur = median(nondur/durables))
```






# more `ddply{plyr}`

* Split data frame, apply function, and return results in a data frame.
```{r,message=FALSE}
library(plyr)
ddply(.data=dat.inc,.variables="sex",.fun=summarise,inc=mean(hhinc))  # mean income by sex?
ddply(dat.inc,c("sex","numads"),summarise,alc=mean(alc),foodin=mean(foodin),foodout=mean(foodout)) 
```

# `ddply{plyr}` 2

* `summarise` can be many things, not just `mean()`
* you can supply any function you want
* check out Hadley's tutorial. ```ddply``` is very powerful.
```{r}
ddply(dat.inc,"sex",summarise,age.range=median(age),car.rooms=max(ncars/nrooms)) 
ddply(subset(dat.inc,numhhkid > 1),"ncars",function(x) data.frame(dur.inc=mean(x$dur/x$hhinc),nondur.inc=mean(x$nondur/x$hhinc)))  # instead of summarise(), supply an "anonymous" function
```

# Alternatives to ddply

* very powerful and elegent package: ```data.table```
* a `data.table` is a `data.frame` on speed
* excellent documentation and explainers (*vignette*) [at CRAN](http://cran.r-project.org/web/packages/data.table/index.html)
* very fast 
* maybe easier to understand than ddply
* read the 10 minutes intro on the CRAN link above

# summary by group with `data.table`

* a data.table has two indices, `i` and `j`: `DT[i,j]`
* `i` is SQL *where* and `j` is an expression

```{r}
library(data.table)
dat.tab <- data.table(dat.inc)
tables()
dat.tab[,list(median.age=as.numeric(median(age)),cars.over.rooms=median(ncars/nrooms)),by=sex]
# adding a column "by reference" with :=
dat.tab[,rel.inc := hhinc/median(hhinc)]  # I typed "rel.inc := hhinc/median(hhinc)" here. 
dat.tab[,rel.inc.sex := hhinc/median(hhinc),by=sex]
dat.tab[,rel.inc.sex.nrooms := hhinc/median(hhinc),by=list(sex,nrooms)]
```

# data.tables and keys

* very large speed gain by avoiding *vector scan* and using *binary search* instead
* you set a *key* on a table, like a database

```{r}
setkey(dat.tab,numhhkid)
dat.tab[numhhkid==1]
dat.tab[J(1)]
dat.tab[J(2),twokid.foodin := log(foodin)/log(foodout)]
```



# Graphics and R

* production of high-quality graphics is another strength of R
* base R is a good start. Many class-specific plotting methods.
* more functionality from other packages:
  * `library(ggplot2)`
  * `library(lattice)`
* one could easily fill 2 hours talking about ggplot alone. Barely scratching the surface here.

# Base R plots  

```{r}
plot(y=log(dat.inc$ndex),x=log(dat.inc$hhinc)) # many options. see ?plot and ?par
plot(factor(dat.inc$ncars)) # plot depends on data types
```

# Plots with ggplot

* ggplots are based on the Grammar of Graphics
* [click here for help, examples and tutorials at http://ggplot2.org/](http://ggplot2.org/)
* a plot is defined by a set of data that is mapped onto a picture - in many complicated ways, if you so desire
* the mappings are called ```aes()``` thetics. For example ```aes(x=var1,y=var2)```
* the way they appear in the picture are called ```geom``` etrics
* there is a very long list of ```geom```s. Refer to documentation.
* let's plot log non-durable expenditure against log household income in a scatter plot
```{r}
library(ggplot2)  
p <- ggplot(data=dat.inc,aes(x=log(hhinc),y=log(ndex)))	# not a plot yet: misses a geom
p + geom_point() # now that's a plot
```

# ggplot 2

```{r gg1,cache=TRUE,fig.width=10}
dat <- subset(dat.inc, hhinc>0 & ndex > 0) # get rid of some negative values
# add another layer: color points by marstat, and decrease opacity to 0.4
p <- ggplot(data=dat,aes(x=log(hhinc),y=log(ndex)))  # not a plot yet: misses a geom
p + geom_point(aes(color=marstat), alpha=0.4) 
p + geom_point() + facet_wrap( ~ sex) # split plot by factor sex
```

# More ggplots

```{r gg2,cache=TRUE,fig.width=10}
ggplot(data=dat,aes(x=log(hhinc),y=log(ndex))) + geom_point(aes(color=age.cat),alpha=0.4)
ggplot(data=dat,aes(x=log(hhinc),y=log(ndex))) + geom_smooth(aes(color=age.cat),method="lm",se=FALSE) 
sdat <- subset(dat,ced.cat %in% c("(10,15]","(15,20]","(20,25]") & log(hhinc) > 3& log(hhinc) < 9) # select a subset of ced
ggplot(data=sdat,aes(x=log(hhinc),y=log(ndex))) + geom_point(aes(color=ced.cat),alpha=0.3) + geom_smooth(aes(color=ced.cat),se=FALSE,size=1)
# split by sex?
ggplot(data=sdat,aes(x=log(hhinc),y=log(ndex))) + geom_point(aes(color=ced.cat),alpha=0.3) + geom_smooth(aes(color=ced.cat),se=FALSE,size=1) + facet_wrap(~sex)
```

# Regression Models

* linear models: ```lm()```
* generalized linear models (probit, logit et al): ```glm```
* loads of packages on [CRAN](http://cran.r-project.org/)
* specify a model with a formula
```{r}
mod.1 <- lm(formula= lndex ~ log(hhinc),data=dat)
summary(mod.1)
```

# ```update()``` a model

* easy to iterate on models
```{r}
mod.2 <- update(mod.1, . ~ . -1 )  # leave formula as is, but substract the intercept
mod.3 <- update(mod.1, . ~ . + age + I(age^2) )  # add age and age squared
coef(summary(mod.2)) # use coef() function to extract coefs only
coef(summary(mod.3)) 
```

# ```predict()``` data from a model

* use the ```predict()``` function 
* most regression functions have got their own prediction method.
* say we want to predict values of lndex from ```mod.3```
```{r}
newdat <- expand.grid(hhinc=c(200,300),age=c(20,50,80))
cbind(newdat,predict(mod.3,newdat))
```

# ```ggplot``` model coeficients

* ```abline()``` is a convenient ```geom``` for this
* works well for single explanatory variable
```{r gg3,cache=TRUE}
p <- ggplot(dat,aes(x=log(hhinc),y=lndex))
p + geom_point() + geom_abline(intercept=coef(mod.1)[1],slope=coef(mod.1)[2],col="red")
```

# ```update()``` a model 2: add a dummy variable

* let's add the number of rooms as a dummy variable `d`
* i.e. we want d1=1 if nrooms==1, d1=0 else; d2=1 iff nrooms==2 etc
* R does that for you (you don't want to create all those d's)
```{r demo18}
mod.4 <- update(mod.3, . ~ . + factor(nrooms) )  
summary(mod.4)
```

# Hyothesis Testing, IV models

* check out ```library(car)``` and ```library(AER)```
* convenient functions for all applied econometrics tasks

# Create an Age Profile for consumption

* How does the distribution of non-durable expenditure vary wity age?
* How to model this in a regression model?
```{r gg4,cache=TRUE,fig.width=10}
ggplot(subset(dat,ndex < 1000), aes(x=ndex)) + geom_density(aes(color=age.cat))
```

# Age Profile: Quantile Regression

```{r,message=FALSE}
library(quantreg)
qreg50 <- rq(formula= ndex ~ age,data=dat,tau=0.5) # median regression
qreg <- rq(formula= ndex ~ age,data=dat,tau=seq(0.05,0.95,le=10)) # quantile reg at several quantiles
plot(summary(qreg)) # very nice plotting method
```

# Age Profile: Compare OLS and Quantile Reg

* compare the Conditional Expectation (the OLS estimate) with the conditional quantile function
* fit a 5-th order polynomial in age
```{r,tidy=FALSE}
form    <- ndex ~ age + I(age^2) + I(age^3) + I(age^4) + I(age^5)
lin.mod <- lm(formula = form, data=dat)
taus    <- seq(0.1,0.9,le=9) # desired quantiles
qr.mod  <- rq(formula= form,data=dat,tau=taus)
ages    <- sort(unique(dat$age)) # vector of ages in data
# create data.frame with cols "age", and a prediction for each model
preds   <- data.frame(age = ages, OLS= predict(lin.mod, newdata = data.frame(age = ages)))
# add qreg predictions
preds   <- cbind(preds, predict(qr.mod,data.frame(age=ages)))
head(preds)
```

# Plot OLS vs Quantile Reg Age Profiles

* ggplot works best with long data. 
* need to ```reshape``` preds into long
```{r gg5,cache=TRUE,message=FALSE,fig.width=10}
library(reshape)
mpreds <- melt(preds,id.vars="age") # now the data is long
head(mpreds)
ggplot(subset(mpreds,age>22),aes(x=age,y=value,color=variable)) + geom_line() + scale_y_continuous(name="predicted nondurable expenditure")
```

# R and Time Series

* R has strong time series capabilities
* base time series architecture in `ts()`
* more advanced ways to deal with time/date classes
    * [Lubridate](http://cran.r-project.org/web/packages/lubridate/index.html)
    * [xts](http://cran.r-project.org/web/packages/xts/index.html)
* check out 
    * [CRAN Time Series Task View: General Overview](http://cran.stat.ucla.edu/web/views/TimeSeries.html)
    * ```library(zoo)``` for general ordered obs
    * ```library(dynlm)``` for dynamic regressions
```{r}
data(UKDriverDeaths)
UK <- UKDriverDeaths
str(UK)
plot(UK)
plot(decompose(UK))
acf(UK)
pacf(UK)
```

# R and panel data

* package [plm](http://cran.r-project.org/web/packages/plm/)
* others...

# R and multinomial logit

* [mlogit package](http://cran.r-project.org/web/packages/mlogit/)

# R and Foreign Languages

* It is easy to call other languages from within R
* R is itself written in Fortran and C, so those two are particularly easy
* This is important for R users who want a lot of performance (MCMC, dynamic programming, etc)

# R and Fortran

* suppose we have the fortran *submodule* `bar.f` stored on disk
* it takes integer `n`, makes `x=1:n` and returns `x^2`
* it looks like that

`````
    subroutine bar(n, x)
        integer n
        double precision x(n)
        integer i
        do 100 i = 1, n
            x(i) = x(i) ** 2 
    100 continue
    end
`````

* we need to compile the function. write in your terminal

`````
cd ~/git/R-demo
R CMD SHLIB bar.f
`````

* It's already compiled in the github repository so no need to if you downloaded that.
* finally, in R we need to load the *dynamic library* and call the function
```{r,tidy=FALSE}
dyn.load("bar.so")
how.many <- 5L # L means integer
y        <- rnorm(how.many)
r.result <- y^2
# ideally we have a wrapper on .Fortran to check sanity of input
f.result <- .Fortran("bar", n=how.many, x=y) # .Fortran returns a list
identical(r.result,f.result$x) #identical?
```

# R and C/C++

* `library(Rcpp)`
* [Rcpp Mastermind Dirk Eddelbuettel](http://dirk.eddelbuettel.com/code/rcpp.html)
* you can write packages the use C code via Rcpp
* very useful if you have a part of your code that needs speeding up.
* many derived libraries
    * `library(RcppArmadillo)`: interface to the Armadillo C++ matrix algebra library. Armadillo is close to matlab.
    * `library(RcppEigen)`: interface to the C++ Eigen libraries. Eigen is a very versatile matrix algebra library. Great for sparse matrices.
    * `library(RcppSimpleTensor)`: [High dimensional computations made easy](https://github.com/tlamadon/RcppSimpleTensor)
    * many more
* you can write C code *inline* and test it

# R and C++ inline

* `library(inline)` allows *inline* testing of C code
```{r,tidy=FALSE}
library(inline)
my.cpp.code <- "NumericVector xx(x); 
                return wrap( std::accumulate( xx.begin(), xx.end(), 0.0));"
my.cpp.fun  <- cxxfunction( signature( x="numeric"), body=my.cpp.code, plugin="Rcpp")
y           <- seq(1,10,0.5)
r.result    <- sum(y)
c.result    <- my.cpp.fun(y)
identical(r.result,c.result)
```

# How to report results

* [knitr](http://yihui.name/knitr/)
* [reproducible research](http://cran.open-source-solution.org/web/views/ReproducibleResearch.html)



# How to make this set of slides

* Use [RStudio](http://www.rstudio.org) to author an R markdown file
* click a button in RStudio to knit an HTML document
* this uses the amazing [knitr package](http://yihui.name/knitr/) to "knit" (evaluated or not) R code into text documents. 
* finally use pandoc to convert to HTML5 slides: ```pandoc -s -S -t slidy --mathjax intro.md -o intro.html```
* Get the R markdown source code of this presentation on [github](https://github.com/floswald/R-demo)
* More info at [MAGES blog](http://lamages.blogspot.co.uk/2012/05/interactive-html-presentation-with-r.html)
